---
title: "Class10 - Halloween Mini Project"
author: "Adam Bisharat"
format: pdf
---

```{r}
candy_file <- "candy-data.csv"
candy <- read.csv(candy_file, row.names = 1)
head(candy)
```


> Q1. How many different candy types are in this dataset?
```{r}
nrow(candy)
```

85 candy types in this data set

> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

There are 38 fruity candies in the dataset.


> Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
candy["Kit Kat",]$winpercent
```
My favorite candy is Kit Kat and its winpercent value is 76.7686


> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat",]$winpercent
```
winpercent value is 76.7686 (That was not planned kit kats are actually my favorite)

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars",]$winpercent
```


```{r}
library(dplyr)
```
```{r}
candy |>
filter(rownames(candy)=="Haribo Happy Cola") |>
select(winpercent)
```

```{r}
candy |>
filter(winpercent > 50) |>
filter(fruity==1)
```


```{r}
candy[candy$winpercent > 50,][candy$fruity==1,]
```

To get a quick insight into a new dataset some folks like using the skimer package and its
skim() function.

```{r}
library("skimr")
skim(candy)
```

```{r}
skimr::skim(candy)
```


> Q6. Is there any variable/column that looks to be on a different scale to the
majority of the other columns in the dataset?

Yes, winpercent

> Q7. What do you think a zero and one represent for the candy$chocolate column?

No chocolate in that column.

> Q8. Plot a histogram of winpercent values

This can be done using the “base” in R hist() function or ggplot()

```{r}
hist(candy$winpercent, breaks=30)
```



```{r}
library(ggplot2)
ggplot(candy) +
aes(winpercent) +
geom_histogram(binwidth=8) +
theme_bw()
```

> Q9. Is the distribution of winpercent values symmetrical?

No, the distribution is not perfectly symmetrical.

> Q10. Is the center of the distribution above or below 50%?

```{r}
summary(candy$winpercent)
```

The median is below 50% (47.8%) so the center of distribution is below 50%.


> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

On average chocolate candy is higher ranked than fruit candy.

```{r}
fruit.candy <- candy |>
filter(fruity==1)
summary(fruit.candy$winpercent)
```


```{r}
choc.candy <- candy |>
filter(chocolate==1)
summary(choc.candy$winpercent)
```


```{r}
summary(candy[as.logical(candy$chocolate),]$winpercent)
```


```{r}
t.test(choc.candy$winpercent, fruit.candy$winpercent)
```
P-Value < 0.05 (p-value = 2.871e-08) so it is statistically signifigant.

> Q13. What are the five least liked candy types in this set?

The 5 least liked candies include Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, and
Jawbusters.

```{r}
play <- c("d", "a", "c")
sort(play)

order(play)

play[order(play)]


```


```{r}
head(candy[order(candy$winpercent), ], 5)
```


> Q14. What are the top 5 all time favorite candy types out of this set?

The top 5 candies are Snickers, Kit Kat, Twix, Reese’s Miniatures, and Reese’s Peanut Butter
cups.


```{r}
tail(candy[order(candy$winpercent), ], 5)
```

> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
library(ggplot2)
ggplot(candy) +
aes(winpercent, rownames(candy)) +
geom_col()
```

> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
library(ggplot2)
ggplot(candy) +
aes(winpercent, reorder(rownames(candy),winpercent)) +
geom_col()
```


```{r}
ggplot(candy) +
aes(x=winpercent, y=reorder(rownames(candy),winpercent),
fill=chocolate) +
geom_col()
```



```{r}
mycols <- rep("black", nrow(candy))
mycols[as.logical(candy$chocolate)] <- "chocolate"
mycols[as.logical(candy$bar)] <- "brown"
mycols[as.logical(candy$fruity)] = "pink"
mycols
```


```{r}
mycols[rownames(candy)=="Twix"] <- "blue"
ggplot(candy) +
aes(x=winpercent, y=reorder(rownames(candy),winpercent),
fill=chocolate) +
geom_col(fill=mycols)
```



> Q17. What is the worst ranked chocolate candy?

The worst ranked chocolate candy is Sixlets.

> Q18. What is the best ranked fruity candy?

The best ranked fruity candy is Starburst.


```{r}
mycols <- rep("black", nrow(candy))
mycols[rownames(candy)=="Twix"] <- "blue"
mycols[as.logical(candy$chocolate)] <- "chocolate"
mycols[as.logical(candy$bar)] <- "brown"
mycols[as.logical(candy$fruity)] = "pink"
mycols[as.logical(candy$fruity)] <- "red"
ggplot(candy) +
aes(winpercent, pricepercent) +
geom_point(col=mycols)
```

```{r}
ggplot(candy) +
aes(winpercent, pricepercent, label=rownames(candy)) +
geom_point(col=mycols) +
geom_label()
```


```{r}
library(ggrepel)
ggplot(candy) +
aes(winpercent, pricepercent, label=rownames(candy)) +
geom_point(col=mycols) +
geom_text_repel(col=mycols, max.overlaps = 5)
```


> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Reece’s miniatures

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```


> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

The 5 most expensive candies in the dataset are: Nik L Nip, Nestle Smarties, Ring pop,
Hershey’s Krackel, and Hershey’s Milk Chocolate. The least popular is Nip L Nip.

```{r}
library(corrplot)

cij <- cor(candy)
cij
```



```{r}
corrplot(cij)
```


> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Fruit and chocolate are anti-correlated.

> Q23. Similarly, what two variables are most positively correlated?

The two most positively correlated are chocolate and winpercent.



```{r}
pca <- prcomp(candy, scale=T)
summary(pca)
```

```{r}
PCA <- plot(pca$x[,1:2], col=mycols, pch=16)
```




```{r}
pca$rotation
```

```{r}
loadings <-as.data.frame(pca$rotation)
ggplot(loadings) +
aes(PC1, reorder(rownames(loadings), PC1), fill=PC1) +
geom_col()
```


```{r}
my_data <- cbind(candy, pca$x[,1:3])
p <- ggplot(my_data) +
aes(x=PC1, y=PC2,
size=winpercent/100,
text=rownames(my_data),
label=rownames(my_data)) +
geom_point(col=mycols)
p
```

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```


> Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

Fruity, pluribus, and hard are picked up strongly by PC1. This makes sense since the
most popular candies are fruity and chocolate candies in bags and boxes.








